---
title: "Why Hinton Diagrams?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why Hinton Diagrams?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  = 5,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(gghinton)
library(ggplot2)
```

Hinton diagrams were introduced by Geoffrey Hinton, one of the founders of
deep learning, as a practical debugging tool for neural network weights in the
1980s. The diagram appeared in textbooks on neural networks and connectionist
models and became a standard visualization in that literature. Despite their
long history, they remain underused in modern data analysis toolkits, in part
because no convenient, ggplot2-native implementation existed.

`gghinton` aims to fix that.

## The problem with heatmaps for signed data

Suppose you are training a neural network and you want to inspect a weight
matrix: to understand which connections are large, which are small, and
which are inhibitory versus excitatory.  The standard tool is a heatmap:

```{r heatmap-example}
set.seed(7)
nr <- 10
nc <- 18
W <- matrix(rnorm(nr*nc, sd = 0.4), nrow = nr, ncol = nc)
rownames(W) <- paste0("neuron_", 1:nr)
colnames(W) <- paste0("input_",  1:nc)

# The standard heatmap approach
df <- as.data.frame(as.table(W))
names(df) <- c("row", "col", "value")

ggplot(df, aes(x = col, y = row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0) +
  coord_fixed() +
  theme_minimal() +
  theme(panel.grid = element_blank(), 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title = "Weight matrix as a heatmap")
```

This works, but it has weaknesses:

1. **Colour choice matters a lot.** Blue/white/red is readable; many diverging
   palettes are not (especially for colourblind readers).
2. **Small differences are hard to judge.** Is `0.35` more than twice `0.16`?
   With colour, you can't easily tell.
3. **Near-zero entries look similar** to each other and to slightly
   positive/negative entries.

Now the same data as a Hinton diagram:

```{r hinton-example}
df_h <- matrix_to_hinton(W,
  rowname_col = "row", colname_col = "col", value_col = "weight")

ggplot(df_h, aes(x = col, y = row, weight = weight)) +
  geom_hinton() +
  scale_fill_hinton() +
  scale_x_continuous(breaks = seq_along(colnames(W)), labels = colnames(W)) +
  scale_y_continuous(breaks = seq_along(rownames(W)), labels = rev(rownames(W))) +
  scale_y_reverse() +
  coord_fixed() +
  theme_hinton() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  labs(title = "Weight matrix as a Hinton diagram")
```

The key differences:

- **Dominant weights are immediately visible**: large squares catch the eye.
- **Near-zero weights are nearly invisible**: the background shows through.
- **Sign is black-and-white**: no colour palette decisions, no colourblind
  concerns.
- **Magnitude comparisons are accurate**: area comparisons are
  pre-attentive and well-calibrated in human vision.

## Why area beats colour for magnitude

A large body of research in visual perception (Mackinlay 1986; Cleveland &
McGill 1984) ranks visual encoding channels by how accurately humans can
decode quantitative information.  The consensus ranking for magnitude:

1. Position on a common scale (best)
2. Length
3. **Area**
4. Angle / slope
5. Colour saturation (worst for quantitative comparison)

Heatmaps use **colour saturation** (the worst channel for magnitude).  Hinton
diagrams use **area** (a dramatically better channel).  The improvement is most
pronounced when:

- Values span a wide range (e.g., `0.01` to `0.99`): tiny vs large squares
  are unmistakable; pale vs saturated blue is not.
- You need to compare non-adjacent entries: spatial position makes area
  comparisons easy across the matrix.

## The signed data advantage

For correlation matrices or weight matrices where sign matters, Hinton
diagrams have an additional advantage.  A heat map must choose a diverging
colour scheme, map its midpoint correctly to zero, and hope that readers
can distinguish near-zero from slightly-positive from slightly-negative.

A Hinton diagram encodes sign with the most basic visual distinction
possible: **black vs white**.  There is no perceptual ambiguity.

```{r correlation-example}
set.seed(3)
# Simulate a correlation matrix
S <- matrix(c(
   1.00,  0.72, -0.35,  0.15,
   0.72,  1.00, -0.21,  0.08,
  -0.35, -0.21,  1.00, -0.58,
   0.15,  0.08, -0.58,  1.00
), 4, 4)
vars <- c("IQ", "Memory", "Anxiety", "Stress")
rownames(S) <- colnames(S) <- vars

df_cor <- matrix_to_hinton(S)

ggplot(df_cor, aes(x = col, y = row, weight = weight)) +
  geom_hinton() +
  scale_fill_hinton() +
  scale_x_continuous(breaks = 1:4, labels = vars) +
  scale_y_continuous(breaks = 1:4, labels = rev(vars)) +
  coord_fixed() +
  theme_hinton() +
  labs(title = "Correlation matrix",
       subtitle = "White = positive, black = negative")
```

Notice how the `Anxiety-Stress` negative correlation is immediately visible as
a large black square, while the small positive `IQ-Stress` correlation
is nearly absent.

## When other visualisations are better

Hinton diagrams are not universally superior. Other visualisations may be better when:

- **The matrix is large** (say, > 50x50). Hinton squares become tiny and the
  visual advantage becomes less clear.
- **You need to communicate exact values**: Hinton diagrams prioritize relative visual impression. Other representations with numeric labels may be a better choice if precision is important.
- **Continuous gradients** matter more than individual entries (e.g., a
  spatial field like temperature over a map).
